# **Caspase-4 Expression Analysis in Mtb-infected Macrophages**

---
title: "Caspase-4 Expression Analysis in Mtb-infected Macrophages"
author: "Ophélie Rutschmann"
source: \>
Data obtained from:
Davide Pisu, Lu Huang, Vipin Narang, Monique Theriault, Gabrielle Lê-Bury,
Bernett Lee, Agnes E. Lakudzala, David T. Mzinza, David V. Mhango,
Steven C. Mitini-Nkhoma, Kondwani C. Jambo, Amit Singhal,
Henry C. Mwandumba, David G. Russell;
Single cell analysis of M. tuberculosis phenotype and macrophage
lineages in the infected lung. J Exp Med 6 September 2021; 218 (9): e20210615.
doi: https://doi.org/10.1084/jem.20210615
date: "September 2023"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6
)
```

## Overview

This analysis examines **Caspase-4 (Also known as Caspase-11)** expression across macrophage populations in a single-cell RNA sequencing dataset of *Mycobacterium tuberculosis* (Mtb)-infected mouse lungs (Original data from Davide Pisu, Lu Huang, Vipin Narang, Monique Theriault, Gabrielle Lê-Bury, Bernett Lee, Agnes E. Lakudzala, David T. Mzinza, David V. Mhango, Steven C. Mitini-Nkhoma, Kondwani C. Jambo, Amit Singhal, Henry C. Mwandumba, David G. Russell; Single cell analysis of M. tuberculosis phenotype and macrophage lineages in the infected lung. J Exp Med 6 September 2021; 218 (9): e20210615. doi: <https://doi.org/10.1084/jem.20210615>). Cells are separated by infection status — **infected**, **bystander**, and **uninfected** — and differential expression is assessed within each macrophage subpopulation.

The analysis is done directly on the GSE167232_mtb_integrated.rds file containing the seurat object obtained from <https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE167232>.

## Libraries

```{r libraries}
library(Seurat)
library(ggplot2)
library(cowplot)
library(clusterProfiler)
library(enrichplot)
library(org.Mm.eg.db)
library(plyr)
library(vioplot)
library(dplyr)
library(patchwork)
library(ggupset)
```

------------------------------------------------------------------------

## Helper Functions

The `initialization()` function preprocesses the Seurat object by:

-   Saving current cluster identities as `Cell_identity`
-   Recoding cluster labels into broader `Cell_type` categories (alveolar macrophage, interstitial macrophage, DC, other)
-   Creating an infection `Status` variable from the `MULTI_ID` field (`infected`, `bystander`, `uninfected`)
-   Adding binary `Casp4_status` (expressing / not_expressing) and continuous `Casp4_value` features

```{r initialization-function, cache=TRUE}
initialization <- function(so) {

  # Save current identities as a new variable
  so[["Cell_identity"]] <- Idents(so)

  # Create Cell_type category
  so[["Cell_type"]] <- so$Cell_identity
  so$Cell_type <- recode(so$Cell_type,
    "AM_1" = "Alveolar_mac", "AM_2" = "Alveolar_mac",
    "AM_3" = "Alveolar_mac", "AM_4" = "Alveolar_mac"
  )
  so$Cell_type <- recode(so$Cell_type,
    "IM_1" = "Interstitial_mac", "IM_2" = "Interstitial_mac",
    "IM_3" = "Interstitial_mac", "IM_4" = "Interstitial_mac"
  )
  so$Cell_type <- recode(so$Cell_type,
    "DC.103+11B-" = "DC",
    "Macrophages/T-cells" = "Other",
    "Macrophages/B-cells" = "Other"
  )

  # Save infection status from MULTI_ID
  so[["Status"]] <- so$MULTI_ID
  so$Status[so$MULTI_ID == "gfp-hi" | so$MULTI_ID == "gfp-low"] <- "infected"
  so$Status <- factor(so$Status, levels = c("uninfected", "bystander", "infected"))

  # Create a new variable with Casp4 expression status
  so[["Casp4_status"]] <- so@assays$RNA@data[
    match("Casp4", so@assays$RNA@data@Dimnames[[1]]), ]
  so$Casp4_status[so$Casp4_status > 0]  <- "expressing"
  so$Casp4_status[so$Casp4_status <= 0] <- "not_expressing"

  # Create a new feature with the Casp4 expression value
  so[["Casp4_value"]] <- so@assays$RNA@data[
    match("Casp4", so@assays$RNA@data@Dimnames[[1]]), ]

  return(so)
}
```

------------------------------------------------------------------------

## Load and Initialize Dataset

```{r load-data, cache=TRUE}
so <- readRDS("GSE167232_mtb_integrated.RDS")
so <- UpdateSeuratObject(so)
so <- initialization(so)
```

```{r define-groups, cache=TRUE}
# Cell population groupings
alveolar     <- c("AM_4", "AM_1", "AM_2", "AM_3")
interstitial <- c("IM_1", "IM_2", "IM_3", "IM_4")
macrophages  <- c(alveolar, interstitial)

# Convenience logical vectors
is_infected  <- as.logical(so$Status == "infected")
is_Casp4pos  <- as.logical(so$Casp4_value == "expressing")
```

```{r colour-palette, cache=TRUE}
# Create a global colour palette (sienna / dark-blue) for the graphs
# Three infection statuses
status_cols <- c(
  uninfected = "sienna4",
  bystander  = "sienna2",
  infected   = "midnightblue"
)

# Eight macrophage subpopulations (AM_1–4 in blues, IM_1–4 in siennas)
mac_cols <- c(
  AM_4       = "midnightblue",
  AM_1       = "dodgerblue4",
  AM_2       = "dodgerblue3",
  AM_3       = "dodgerblue2",
  IM_1       = "skyblue3",
  IM_2       = "sienna4",
  IM_3       = "sienna3",
  IM_4       = "sienna2"
)

# Extended palette for DimPlot (all cell identities)
cell_identity_cols <- c(
  "midnightblue", "dodgerblue4", "dodgerblue3", "dodgerblue2",
  "skyblue4",     "skyblue3",    "sienna4",     "sienna3",
  "sienna2",      "sienna1",     "tan4",        "tan3",
  "tan2",         "tan1"
)

# Continuous gradient endpoints
col_low  <- "sienna4"
col_high <- "midnightblue"
```

------------------------------------------------------------------------

## Alveolar Macrophage Subpopulations

### AM_1

```{r AM1, cache=TRUE}
Idents(so) <- "Cell_identity"
AM1_so <- subset(so, idents = "AM_1")
Idents(AM1_so) <- "Status"

AM1.de.markers <- FindMarkers(AM1_so, ident.1 = "infected", ident.2 = "uninfected", only.pos = TRUE)
AM1.de.markers$gene <- rownames(AM1.de.markers)

VlnPlot(AM1_so, features = "Casp4", cols = status_cols)
```

> **Summary:** Casp4 is upregulated in infected vs. uninfected AM_1 cells, but not in bystanders.

------------------------------------------------------------------------

### AM_2

```{r AM2, cache=TRUE}
AM2_so <- subset(so, idents = "AM_2")
Idents(AM2_so) <- "Status"

AM2.de.markers <- FindMarkers(AM2_so, ident.1 = "bystander", ident.2 = "uninfected", only.pos = TRUE)
AM2.de.markers$gene <- rownames(AM2.de.markers)

VlnPlot(AM2_so, features = "Casp4", cols = status_cols)
```

> **Summary:** Casp4 is not upregulated in AM_2.

------------------------------------------------------------------------

### AM_3

```{r AM3, cache=TRUE}
AM3_so <- subset(so, idents = "AM_3")
Idents(AM3_so) <- "Status"

AM3.de.markers <- FindMarkers(AM3_so, ident.1 = "infected", ident.2 = "uninfected", only.pos = TRUE)
AM3.de.markers$gene <- rownames(AM3.de.markers)

VlnPlot(AM3_so, features = "Casp4", cols = status_cols)
```

> **Summary:** Casp4 is not upregulated in AM_3.

------------------------------------------------------------------------

### AM_4

```{r AM4, cache=TRUE}
AM4_so <- subset(so, idents = "AM_4")
Idents(AM4_so) <- "Status"

AM4.de.markers <- FindMarkers(AM4_so, ident.1 = "infected", ident.2 = "uninfected", only.pos = TRUE)
AM4.de.markers$gene <- rownames(AM4.de.markers)

VlnPlot(AM4_so, features = "Casp4", cols = status_cols)
```

> **Summary:** Casp4 is not upregulated in AM_4.

------------------------------------------------------------------------

### All Alveolar Macrophages (Grouped)

```{r alveolar-grouped, cache=TRUE}
Alveolar_so <- subset(so, idents = alveolar)
Idents(Alveolar_so) <- "Status"

Alveolar.de.markers <- FindMarkers(Alveolar_so, ident.1 = "infected", ident.2 = "bystander", only.pos = TRUE)
Alveolar.de.markers$gene <- rownames(Alveolar.de.markers)

VlnPlot(Alveolar_so, features = "Casp4", cols = status_cols)

DotPlot(Alveolar_so, features = c("Casp4", "Il1b", "Casp1", "Nlrp3"),
        col.min = 0, cols = c(col_low, col_high)) +
  ggtitle("Alveolar Macrophages")
```

------------------------------------------------------------------------

## Interstitial Macrophage Subpopulations

### IM_1

```{r IM1, cache=TRUE}
Idents(so) <- "Cell_identity"
IM1_so <- subset(so, idents = "IM_1")
Idents(IM1_so) <- "Status"

IM1.de.markers <- FindMarkers(IM1_so, ident.1 = "infected", ident.2 = "uninfected", only.pos = TRUE)
IM1.de.markers$gene <- rownames(IM1.de.markers)

VlnPlot(IM1_so, features = "Casp4", cols = status_cols)
```

> **Summary:** Casp4 is not upregulated in IM_1.

------------------------------------------------------------------------

### IM_2

```{r IM2, cache=TRUE}
IM2_so <- subset(so, idents = "IM_2")
Idents(IM2_so) <- "Status"

IM2.de.markers <- FindMarkers(IM2_so, ident.1 = "bystander", ident.2 = "uninfected", only.pos = TRUE)
IM2.de.markers$gene <- rownames(IM2.de.markers)

VlnPlot(IM2_so, features = "Casp4", cols = status_cols)
```

> **Summary:** Casp4 is not upregulated in IM_2.

------------------------------------------------------------------------

### IM_3

```{r IM3, cache=TRUE}
IM3_so <- subset(so, idents = "IM_3")
Idents(IM3_so) <- "Status"

IM3.de.markers <- FindMarkers(IM3_so, ident.1 = "bystander", ident.2 = "infected", only.pos = TRUE)
IM3.de.markers$gene <- rownames(IM3.de.markers)

VlnPlot(IM3_so, features = "Casp4", cols = status_cols)
DotPlot(IM3_so, features = "Casp4", col.min = 0, cols = c(col_low, col_high))
```

> **Note:** Insufficient uninfected cells for a full differential analysis. DotPlot suggests Casp4 expression is present in this population.

------------------------------------------------------------------------

### IM_4

```{r IM4, cache=TRUE}
IM4_so <- subset(so, idents = "IM_4")
Idents(IM4_so) <- "Status"

IM4.de.markers <- FindMarkers(IM4_so, ident.1 = "infected", ident.2 = "uninfected", only.pos = TRUE)
IM4.de.markers$gene <- rownames(IM4.de.markers)

VlnPlot(IM4_so, features = "Casp4", cols = status_cols)
DotPlot(IM4_so, features = "Casp4", col.min = 0, cols = c(col_low, col_high))
```

------------------------------------------------------------------------

### All Interstitial Macrophages (Grouped)

```{r interstitial-grouped, cache=TRUE}
Interstitial_so <- subset(so, idents = interstitial)
Idents(Interstitial_so) <- "Status"

Interstitial.de.markers <- FindMarkers(Interstitial_so, ident.1 = "bystander", ident.2 = "uninfected", only.pos = TRUE)
Interstitial.de.markers$gene <- rownames(Interstitial.de.markers)

VlnPlot(Interstitial_so, features = "Casp4", cols = status_cols)

DotPlot(Interstitial_so, features = c("Casp4", "Il1b", "Casp1", "Nlrp3"),
        col.min = 0, cols = c(col_low, col_high)) +
  ggtitle("Interstitial Macrophages")
```

------------------------------------------------------------------------

## Other Myeloid Populations

### Neutrophils

```{r neutrophils, cache=TRUE}
N_so <- subset(so, idents = "Neutrophils")
Idents(N_so) <- "Status"

N.de.markers <- FindMarkers(N_so, ident.1 = "infected", ident.2 = "uninfected", only.pos = TRUE)
N.de.markers$gene <- rownames(N.de.markers)

VlnPlot(N_so, features = "Casp4", cols = status_cols)
```

> **Summary:** Casp4 is upregulated in both bystander and infected neutrophils.

------------------------------------------------------------------------

### Monocytes

```{r monocytes, cache=TRUE}
M_so <- subset(so, idents = "Monocytes")
Idents(M_so) <- "Status"

M.de.markers <- FindMarkers(M_so, ident.1 = "bystander", ident.2 = "uninfected", only.pos = TRUE)
M.de.markers$gene <- rownames(M.de.markers)

VlnPlot(M_so, features = "Casp4", cols = status_cols)
```

> **Summary:** Casp4 is upregulated in both bystander and infected monocytes.

------------------------------------------------------------------------

### Dendritic Cells

```{r DCs, cache=TRUE}
DC_so <- subset(so, idents = c("DC", "DC.103+11B-"))
Idents(DC_so) <- "Status"

DC.de.markers <- FindMarkers(DC_so, ident.1 = "infected", ident.2 = "uninfected", only.pos = TRUE)
DC.de.markers$gene <- rownames(DC.de.markers)

VlnPlot(DC_so, features = "Casp4", cols = status_cols)
```

> **Summary:** No Casp4 upregulation detected in dendritic cells.

------------------------------------------------------------------------

## Casp4+ vs. Casp4− Cells in Infected Interstitial Macrophages

This section compares gene expression between Casp4-expressing and non-expressing cells within the **infected** interstitial macrophage compartment, followed by pathway enrichment analysis.

### Differential Expression

```{r casp4-DE, cache=TRUE}
Idents(Interstitial_so) <- "Status"
infected_so <- subset(Interstitial_so, idents = "infected")
Idents(infected_so) <- "Casp4_status"

Casp4.de.markers <- FindMarkers(infected_so, ident.1 = "expressing", ident.2 = "not_expressing")
Casp4.de.markers.pos <- subset(Casp4.de.markers, p_val_adj < 0.1)
Casp4.de.markers.pos$gene <- rownames(Casp4.de.markers.pos)

head(Casp4.de.markers.pos, 20)
```

### Gene Set Enrichment — GO Biological Process

```{r GSEA-GO, cache=TRUE}
de.genes  <- rownames(Casp4.de.markers.pos)
all.genes <- unlist(infected_so@assays$RNA@data@Dimnames[1])

# Keep only genes expressed in at least one cell to reduce memory usage
expressed_mask <- Matrix::rowSums(infected_so@assays$RNA@data > 0) > 0
all.genes <- all.genes[expressed_mask]

# Convert gene symbols to Entrez IDs
de.genes.entrez  <- na.omit(mapIds(org.Mm.eg.db, de.genes,  'ENTREZID', 'SYMBOL'))
all.genes.entrez <- na.omit(mapIds(org.Mm.eg.db, all.genes, 'ENTREZID', 'SYMBOL'))

enrichment.interstitial <- enrichGO(
  gene          = de.genes.entrez,
  ont           = "BP",
  OrgDb         = org.Mm.eg.db,
  universe      = all.genes.entrez,
  readable      = TRUE,
  pvalueCutoff  = 0.05
)

tab.go <- as.data.frame(enrichment.interstitial)
tab.go[1:20, 1:6]
```

### Gene Set Enrichment — KEGG

```{r GSEA-KEGG}
enrichment.kegg <- enrichKEGG(
  gene         = de.genes.entrez,
  organism     = 'mmu',
  universe     = all.genes.entrez,
  pvalueCutoff = 0.05
)

tab.kegg <- as.data.frame(enrichment.kegg)
tab.kegg <- subset(tab.kegg, Count > 0)
tab.kegg[1:20, 1:6]
```

### Enrichment Visualizations

```{r enrichment-plots, fig.height=8}
group_color <- c("midnightblue", "dodgerblue4", "dodgerblue3", "skyblue3", "sienna4", "sienna3", "sienna2")

edox  <- setReadable(enrichment.interstitial, 'org.Mm.eg.db', 'ENTREZID')
edox2 <- pairwise_termsim(edox)

# Tree plot
treeplot(edox2, nCluster = 7, group_color = group_color) +
  scale_color_gradient(low = col_low, high = "sienna1")

# Network plots
cnetplot(edox, foldChange = de.genes.entrez, showCategory = 20,
         color_category = "midnightblue", color_gene = col_low)
goplot(enrichment.interstitial, showCategory = 10)
emapplot(edox2, showCategory = 15)

# Bar and dot plots
barplot(enrichment.interstitial, showCategory = 15) +
  scale_fill_gradient(low = col_low, high = col_high) +
  ggtitle("GO Barplot")
barplot(enrichment.kegg, showCategory = 15) +
  scale_fill_gradient(low = col_low, high = col_high) +
  ggtitle("KEGG Barplot")

dotplot(enrichment.interstitial, showCategory = 16) +
  scale_color_gradient(low = col_low, high = col_high) +
  ggtitle("GO Dotplot")
dotplot(enrichment.kegg, showCategory = 16) +
  scale_color_gradient(low = col_low, high = col_high) +
  ggtitle("KEGG Dotplot")
```

------------------------------------------------------------------------

## Summary Plots

### Violin Plots

```{r violin-summary}
v1 <- VlnPlot(AM1_so,          features = c("Casp4", "Casp1"), cols = status_cols)
v2 <- VlnPlot(Interstitial_so, features = c("Casp4", "Il1b"),  cols = status_cols)
v1 + v2
```

### Dot Plots — Inflammasome Markers

```{r dot-summary}
d1 <- DotPlot(Alveolar_so,     features = c("Casp4", "Il1b", "Casp1", "Nlrp3"),
              col.min = 0, cols = c(col_low, col_high)) +
  ggtitle("Alveolar Macrophages")
d2 <- DotPlot(Interstitial_so, features = c("Casp4", "Il1b", "Casp1", "Nlrp3"),
              col.min = 0, cols = c(col_low, col_high)) +
  ggtitle("Interstitial Macrophages")
d1 + d2
```

### Feature & Scatter Plots

```{r feature-plots}
FeaturePlot(so, features = "Gbp7", split.by = "Status",
            cols = c("lightgrey", col_high))

FeatureScatter(so, "Gbp2", "Casp4", group.by = "Status",
               cols = unname(status_cols))
```

------------------------------------------------------------------------

## Percentage & Distribution Analysis

This section quantifies the proportion of Casp4-expressing cells across macrophage subpopulations and examines how Casp4+ cells are distributed by cell type and infection status.

### Setup

```{r pda-setup}
colorsName <- c("midnightblue", "dodgerblue4", "dodgerblue3", "dodgerblue2",
                "sienna4",     "sienna3",     "sienna2",     "sienna1")

# Infection status logical vectors (full Seurat object)
is_infected   <- as.logical(so$Status == "infected")
is_bystander  <- as.logical(so$Status == "bystander")
is_uninfected <- as.logical(so$Status == "uninfected")

# Casp4 expression logical vector
is_Casp4pos <- as.logical(so$Casp4_status == "expressing")
```

### Casp4+ Frequency in Infected Cells (All Populations)

```{r pda-total-frequency}
# Total frequency of Casp4+ cells among all infected cells
df.infected <- as.data.frame(so$Casp4_status[is_infected])
tot_frequency_infected <- count(df.infected)
tot_frequency_infected
```

### Casp4+ Percentage per Macrophage Subpopulation

For each macrophage subpopulation, the proportion of Casp4-expressing cells among **infected** cells is calculated, alongside the mean Casp4 expression value in **uninfected** cells.

```{r pda-percentages-loop}
percentages <- c()
frequencies <- c()
mean_casp4  <- c()

for (t in macrophages) {
  frequencies <- dplyr::count(
    as.data.frame(so$Casp4_status[is_infected & so$Cell_identity == t]),
    so$Casp4_status[is_infected & so$Cell_identity == t]
  )
  percentages <- c(percentages,
    frequencies[1, 2] / (frequencies[1, 2] + frequencies[2, 2]))
  mean_casp4  <- c(mean_casp4,
    mean(so$Casp4_value[is_uninfected & so$Cell_identity == t]))
  print(t)
}

# Compile results
percentages_perCellPop <- data.frame(Cell_type = macrophages, Percent_pos = percentages)
row.names(percentages_perCellPop) <- macrophages

mean_df <- data.frame(Cell_type = macrophages, Casp4 = mean_casp4)
```

```{r pda-percentages-plot}
ggplot(percentages_perCellPop, aes(x = Cell_type, y = Percent_pos)) +
  geom_bar(stat = "identity", color = "black", fill = colorsName) +
  theme_minimal() +
  labs(title = "Percentage of Casp4+ Cells per Macrophage Subpopulation (Infected)",
       x = "Cell Type", y = "Proportion Casp4+")
```

### Distribution of Casp4+ Cells by Cell Type

This analysis asks: among Casp4-expressing **bystander** macrophages, what is the cell-type composition?

```{r pda-distribution}
# Subset to macrophage populations only
Idents(so) <- "Cell_identity"
macrophages_so <- subset(so, idents = macrophages)

# Redefine status vectors for the macrophage subset
is_infected   <- as.logical(macrophages_so$Status == "infected")
is_uninfected <- as.logical(macrophages_so$Status == "uninfected")
is_bystander  <- as.logical(macrophages_so$Status == "bystander")
is_Casp4pos   <- as.logical(macrophages_so$Casp4_status == "expressing")

# Frequency of each cell type among Casp4+ bystander macrophages
frequency_of_each_cell_type <- dplyr::count(
  as.data.frame(macrophages_so$Cell_identity[is_Casp4pos & is_bystander]),
  macrophages_so$Cell_identity[is_Casp4pos & is_bystander]
)

# Normalise to proportions
percentage_of_each_cell_type <- frequency_of_each_cell_type
percentage_of_each_cell_type[, 2] <- percentage_of_each_cell_type[, 2] /
  length(so@assays$RNA@data@p[is_Casp4pos & is_bystander])
```

### Pie Chart: Casp4+ Bystander Macrophages

```{r pda-pie}
pie(percentage_of_each_cell_type[, 2],
    labels   = percentage_of_each_cell_type[, 1],
    clockwise = TRUE,
    main     = "Casp4+ Bystander Macrophages",
    col      = colorsName)
```

### Bar Plot: Percentage of Casp4+ Cells

```{r pda-barplot}
barplot(percentages_perCellPop[, 2],
        col      = colorsName,
        names.arg = percentages_perCellPop[, 1],
        ylim     = c(0, 0.40),
        main     = "Percentage of Casp4+ Cells per Subpopulation",
        las      = 2,
        ylab     = "Proportion Casp4+")
```

### Split Violin Plots: Casp4 Expression by Status

```{r pda-split-violin}
casp4_inf   <- as.data.frame(c())
casp4_bys   <- as.data.frame(c())
casp4_uninf <- as.data.frame(c())

for (identity in macrophages) {
  casp4_inf   <- c(casp4_inf,   as.data.frame(so$Casp4_value[is_infected   & so$Cell_identity == identity]))
  casp4_bys   <- c(casp4_bys,   as.data.frame(so$Casp4_value[is_bystander  & so$Cell_identity == identity]))
  casp4_uninf <- c(casp4_uninf, as.data.frame(so$Casp4_value[is_uninfected & so$Cell_identity == identity]))
  print(identity)
}

vioplot(casp4_inf,
        plotCentre = "line",
        side       = "right",
        col        = "midnightblue",
        ylim       = c(2, 8),
        wex        = 3,
        names      = c("AM1", "AM2", "AM3", "AM4", "IM1", "IM2", "IM3", "IM4"))

vioplot(casp4_bys,
        plotCentre = "line",
        side       = "left",
        col        = "sienna2",
        add        = TRUE,
        wex        = 3)

vioplot(casp4_uninf,
        plotCentre = "line",
        side       = "right",
        col        = "sienna4",
        add        = TRUE,
        wex        = 3)

legend("topright",
       legend = c("Infected", "Bystander", "Uninfected"),
       fill   = c("midnightblue", "sienna2", "sienna4"),
       bty    = "n")
```

### UMAP & Feature Plots

```{r pda-dimplots}
Idents(macrophages_so) <- "Status"
Casp4_so <- subset(so, idents = "expressing")

Idents(Casp4_so) <- "Cell_identity"
Idents(so)       <- "Cell_identity"

# UMAP split by infection status — all cells
DimPlot(so, split.by = "Status")

# UMAP — Casp4+ cells only, split by status
DimPlot(Casp4_so, split.by = "Status")
```

```{r pda-feature-plots}
# Casp4 expression on UMAP, split by infection status
FeaturePlot(so,
            features = "Casp4",
            cols     = c("#D3D3D315", col_high),
            split.by = "Status",
            pt.size  = 0.3) +
  theme(legend.position = "right")

# Full UMAP coloured by cell identity
DimPlot(so,
        group.by = "Cell_identity",
        cols     = cell_identity_cols)
```

### Dot Plots: Casp4 by Cell Identity and Infection Status

```{r pda-dotplots}
Idents(so) <- "Status"

d1 <- DotPlot(macrophages_so, features = "Casp4", group.by = "Cell_identity",
              idents = "uninfected", col.min = 0, cols = c(col_low, col_high)) +
  ggtitle("Uninfected")

d2 <- DotPlot(macrophages_so, features = "Casp4", group.by = "Cell_identity",
              idents = "bystander",  col.min = 0, cols = c(col_low, col_high)) +
  ggtitle("Bystander")

d3 <- DotPlot(macrophages_so, features = "Casp4", group.by = "Cell_identity",
              idents = "infected",   col.min = 0, cols = c(col_low, col_high)) +
  ggtitle("Infected")

d1 + d2 + d3
```

------------------------------------------------------------------------

## Session Info

```{r session-info}
sessionInfo()
```
